A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-16 tex/BiSig.tex
[ tex ] abstract

\begin{abstract}
  Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
  While progress has been made in developing techniques for typechecking type systems on a case-by-case basis, the lack of a general theory poses a difficulty of achieving a `type checker generator' for diverse languages.

  Motivated by this issue, this paper develops a \emph{simple} but \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing for synthesising the type of a program with type annotations.
  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, similar to the role of grammar in parsing.
  Second, our theory is constructive because it is formalised in \Agda and formulated positively in order to compute.
  Specifically, for any mode-correct signature, logical decidability of type synthesis is proved and its proof is instantiated to a synthesiser specific to the corresponding simple type theory, similar to a parser generator.
  The notion of bidirectionalisation is introduced to refine annotability in bidirectional typing and it leads to the trichotomy of raw terms which models a practical type synthesiser that triggers an exception when its input has missing annotations.
%  By formalizing our theory in \Agda,
\end{abstract}

>>>>

\begin{abstract}
  Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
  While progress has been made in developing techniques for typechecking on a case-by-case basis, the lack of a general theory poses a difficulty of implementing a robust `type checker generator' for diverse languages.

  Motivated by this issue, we develop a \emph{simple} but \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a program with type annotations.
  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, a type synthesiser for a simple type theory specified by some mode-correct signature is the proof of logical decidability of synthesis, analogous to a parser generator.
  The notion of bidirectionalisation is introduced to refine annotability in bidirectional typing.
  Type synthesis and bidirectionalisation lead to the trichotomy of raw terms which models a practical type synthesiser that throws an exception when its input does not have enough annotations potentially with a proof to locate missing annotations.
  Hence, by formalizing our general theory of bidirectional type synthesis constructively, we provide a provably correct `type checker generator' for simple type theories, setting the stage for further advances towards a more general theory of type synthesis.
\end{abstract}
